Case-based Arabic Morphological Analysis

### 2010-12-10 tex/rp.tex
modified the Future work section in the proposal, but it needs some proofreading.

@@

\subsection{Budget}

@@

We anticipate that the research will lead to the production of 
CATARM, an Arabic text analysis framework that allows relational 
queries between Arabic text documents.
The components of the framework are listed and discussed
in details in Section~\ref{s:design}.

The following is a list of the anticipated components They are and are repeated for convenience as follows.
\begin{enumerate}
\item Case-based morphological analyzer 
\item Linguistic computational model for Arabic
\item User query language
\item Optimization methodologies
\begin{enumerate}
\item Use of language specific features 
\item Use of data in question specific features
\end{enumerate}
\item Entity and relational extractors
\item Statistical tools
\item Graph and text algorithms
\item Query solver
\item Case studies 
\begin{enumerate}
\item Hadith literature
\item Security
\item Health
\item Network content
\end{enumerate}
\end{enumerate}

\subsection{Evaluation Criteria}

The Lead PI will conduct biweekly meetings to track progress on the project.
 The team will maintain a dokuwiki based website to continuously post updates on the progress of the project.
 In Section "Research Design and Methods" we presented a schedule for delivering the different products of the project.
 The schedule can be used as a timeline for measuring the progress of the project.

The lexical and syntax analysis libraries will be compared with current open source libraries against commonly used corpora of data such as a collection of books or a collection of news reports.
 The dictionary module will be an augmentation over existing digitized dictionaries such as the Buckwalter~\cite{Tim04} morphological analyzer.
 In the entity and relational extraction tools we will use techniques native to the Arabic language and we will compare against English entity and relational extraction tools of similar or translated texts.
 We will also compare techniques that pre-process Arabic and then pass it to English IE and RE techniques.
 The logic solver will build over the IE and RE tools and will use off-the shelf quantifier and propositional solvers such as MiniSat~\cite{Een03}.
 The tagger is a GUI interface that allows fast manual inspection and refinement of the automated results.
 Each tool in the framework can start on its own with a prototypical implementation of the other needed tools.

Each tool in the framework stands alone and can operate on a well defined input and generate the desired output.
 The input of one tool, such as the entity extraction, can be automatically generated by another tool, such as the dictionary and the syntax analyzer.
  However, the input can also be manually generated if desired using the straightforward GUI tagger facility.
 Thus the tagger facility stands as a desired utility on its own and at the same time represents a contingency plan in case any module in the framework failed to deliver.

The team will publish the results of the research in the form of refereed conference and journal papers as well as tool papers addressing each of the tools.
 The team will also hold workshops to educate possible users on the tools once the tools are available.

@@

>>>>

The team will be formed by the two main PIs, Dr. Fadi Zaraket (LPI)
and Dr. Rehab Duwairi (Co-LPI) and three research assistants.
Currently the LPI is working with one research assistant 
and two senior students on an Arabic morphological analyzer.

The LPI offers a course on computational Arabic attended
by excellent senior and graduate students at AUB. 
Once funded, the team can be directly formed and work can start
with productive team members.

@@

The Lead PI will conduct biweekly meetings to track progress 
on the project.
The team will maintain a dokuwiki based website to continuously 
post updates on the progress of the project.

\subsection{Budget items}

In the following we list the budget items. 
Monatary values are provided in more details on the QNRF 
submission website.

\begin{itemize}
\item One research assistant at QU
\item Two research assistants at AUB, one of them will 
work partially from QU. 
\item Two month salary per year for the Co-LPI
\item Two month salary per year for the LPI
\item One course buy out per year for the Co-LPI
\item Two weeks consultant for the second and the third years
\item Licensing and membership fees to acquire digital libraries and corpora
\item High performance computer and storage devices
\item One conference for each PI per year
\item One conference for one RA from QU per year
\item One conference for one RA from AUB per year
\item Site visit to AUB 
\item Site visit to QU 
\item Other miscellanous items
\end{itemize}

@@

We anticipate that the research will lead to the production of 
CATARM, an Arabic text analysis framework that allows relational 
queries between Arabic text documents.
The components of the framework are listed and discussed
in details in Section~\ref{s:design}. 
In the following we will discuss what to anticipate and 
how to evaluate the different components of CATARM.

We will deliver a case-based morphological analyzer. 
The analyzer will be compared to other available analyzers such as
Buckwalter~\cite{Tim04} and ElixirFM~\cite{Otakar:07} in terms
of accuracy and efficiency of analysis.
We will adopt two comparison methodologies. 
First, we will compare the analyzers in an exhaustive analysis mode
which is the only way current analyzers perform. 
Second, we will compare the analyzers under several case-studies
and show how a case-based morphological analyzer can benefit
from case-based optimizations.

\subsection{Linguistic computational model for Arabic}

We will build a linguistic computational model for Arabic 
based on an understanding of the formal Arabic grammar 
and on an abstraction of the current existing concatenation
based lexicons.
We will compare our LCM with that of ElixirGM~\cite{Otakar:07}.
We will study how much such a model can affect the analysis
in terms of accuracy and efficiency as well.

\subsection{User query language}

We will build a query language for relational Arabic text mining 
and a front-end to compile a user query into an optimized
computation. 
We will check how user friendly and how expressive the language is.
We will compare our query language with the sophisticated 
interface of the MADA+TOKAN toolkit~\cite{Habash:09}.
We will look for other attempts to formalize Arabic text 
mining queries and we will also compare agains 
formal text mining queries from other text mining domains such
as the gene ontolgy domain~\cite{GeneOntology10},

\subsection{Optimizations}

We will develop and evaluate our two optimization methodologies
and look for other optimizations. 
First we will use language specific features based on the 
LCM of Arabic, the grammar of the query language, and 
the Arabic query provided by the user to simplify 
the LCM and the morphological analysis. 
Second we will look at language independent techniques
such as HMM and LSI models extracted from the 
text documents and use that to simplify the analysis. 
We will perform the analysis with and without the optimizations
and compare the results in terms of accuracy and efficiency.

%\subsection{ Entity and relational extractors}
\subsection{CATARM and case studies }

We will evaluate CATARM against at least three of the case studies
that we discussed previously and that are listed below. 
\begin{enumerate}
\item Hadith literature
\item Security
\item Health
\item Network content
\end{enumerate}

We anticipate CATARM to be able to successfully solve all the case
studies with resonable computational resources available to the 
average user. 
We will also generalize the components we will develop to solve
the case studies into a library of components that can be used
in other case studies.

\subsection{CATRAM evaluation }

The team will seek the assistance and help of an independent 
consultant in the middle of the second and third years of 
the duration of the project to evaluate and validate the results.

@@

\subsection{Future work}

Completion of the above goals would result in a significant contribution to the research community of NLP in general and Arabic Analysis in particular. However, as already noted, text mining is a wide problem that cannot be expected to saturate in the near future especially when it comes to the arabic language where much work is lacking. Consequently, we look forward for additional contributions to the field such as the following:
\begin{itemize}
\item Enhance morphological analysis: It is always the case, that the lexicon upon which the morphological analyzer can be augmented with useful categorization and entity tags that reflect semantical meanings. These modifications would give CATARM additional capabilities in terms of allowing users to extract and use a greater variety of entities when building instructive queries. The feedback of experts in the field of arabic linguistics can continually add accuracy to the morphological solutions.  
\item Look for other optimizations: From our priliminery research and results we were able to envision some optimizations to current tools; that's why it is expected that as we develop the complete framework to stumble at much more ideas and suggestions for improving the text mining process we are conducting expecially in terms of efficency and accuracy.
\item Generalize the framework to other Arabic-like languages: Some sematic languages share many common features to the Arabic language such as the Amharic, Hebrew, etc... or even are written using the Arabic script like Persian, Kurdish, Urdu, Pashto. Accordingly, one may consider porting our techniques to them, or even generalizing the framework to support for languages other than Arabic.
\item Generalize the case-based approach: ??
\item Integrate other text mining techniques such as vocalization. Vocalization refers to the process of adding diacritics to an unvocalized text. The prerequisites of a successful vocalizer are the accuracy of the morphological analyzer in addition to grammatical analyzers. Such tools are especially useful in the academic sector where fully vocalized texts may be of high utility. 
\item Maintenance: We should never forget that for any software to remain useful, it must be continuously monitored to correct bugs, improve its interface, and enrich its capabilities. That's why CATARM should undergo extensive testing, fixes, and enhancements even after launching. 
\end{itemize}
